Subset coercions in \Coq

### 2007-01-23 papers/article/main.tex
Some more remarks rework

\subsubsection{Properties}
We have proved some of the metatheory for this system in \Coq{}
\cite{sozeau:Coq/Russell/meta-web}.
We have assumed that it is strongly
normalising (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR does not depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalisation could also be proved, perhaps more easily,
using the generic proof method developed by Geuvers \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

@@

\section{Implementation}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The prototype also contains support for strucural and 
well-founded \verb|Fixpoint| definitions.
The type-checker transforms pattern-matching so
that an equality between the matched term and the pattern is present in
the typing context of each branch and it allows the user to put explicit 
holes in the term, for example:
\begin{code}
  Program Definition hd ( l : list A | l <> [] ) : A :=
    match l with
    | hd :: tl -> hd
    | [] -> _
    end.
\end{code}
Here the second branch can't be taken because we would have $l = []$ and
$l \neq []$, but we may not have any placeholder element of type
$A$.

@@

\subsection{Related work}
Other efforts to build certified programs using type theory include
\Cayenne \cite{Augustsson99} and \Epigram
\cite{DBLP:journals/jfp/McBrideM04}. \Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker; it can be thought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
\Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, \Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementally to get a complete program.
Using type annotations and with the help of the editor for structuring, 
one is able to write programs with precise specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scaling issues. The \DML language \cite{Xi98:PhD} is more akin to
comparison with our solution. It provides a way to use a restricted set
of dependent types in \ML programs, carefully chosen so as to keep
type-checking decidable with the help of an automatic prover. Our method
should subsume this one as it is perfectly possible to integrate automatic
proof tools with \Coq to discharge the generated obligations.

>>>>

\subsubsection{Properties}
We have proved some of the metatheory for this system in \Coq{}
\cite{sozeau:Coq/Russell/meta-web}.
We have assumed that it is strongly
normalising (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR does not depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
% Strong normalisation could also be proved, perhaps more easily,
% using the generic proof method developed by Geuvers \cite{geuvers95}, by finding
% a proper interpretation of \Russell terms into the Calculus of
% Constructions.

@@

\section{Implementation}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples (some are available on the website). We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The prototype also contains support for strucural and 
well-founded \verb|Fixpoint| definitions.
The type-checker transforms pattern-matching so
that an equality between the matched term and the pattern is present in
the typing context of each branch and it allows the user to put explicit 
holes in the term, for example:
\begin{code}
  Program Definition hd ( l : list A | l <> [] ) : A :=
    match l with
    | hd :: tl -> hd
    | [] -> False_rec A _
    end.
\end{code}
In the second branch we have $l = []$ and $l \neq []$, hence we can
prove $`_$ and derive an object of type $A$.

@@

\subsection{Related work}
Other efforts to build certified programs using type theory include
\Cayenne \cite{Augustsson99} and \Epigram
\cite{DBLP:journals/jfp/McBrideM04}. \Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker; it can be thought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
\Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, \Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementally to get a complete program.
Using type annotations and with the help of the editor for structuring, 
one is able to write programs with precise specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scaling issues, time will tell which approach is
more beneficial for writing and reasonning about dependently typed
programs. The \DML language \cite{Xi98:PhD} is more akin to
comparison with our solution. It provides a way to use a restricted set
of dependent types in \ML programs, carefully chosen so as to keep
type-checking decidable with the help of an automatic prover. Our method
should subsume this one as it is perfectly possible to integrate automatic
proof tools with \Coq to discharge the generated obligations. There
exists many more systems like Xi's Applied Type Systems and Tim Sheard's
Omega language which explore the design space of dependently typed languages.
